Chess EDA

### 2023-03-22 owen/01_eda.qmd
initial write up

knitr::opts_chunk$set(echo = FALSE)

@@

Now we want to fit this multilevel logistic regression. Letting
$y_{ij}$ be the result of the $i$-th game by the $j$-th focal player,
then the model we will consider is.

---

@@

Instead account for the covariates of colour and elo difference also.

@@

We can then instead include partial pooling of the $\beta$ coefficients.

@@

## Repeat using normalized win Ratio

@@

mcmc_hist(fit3$draws(c("beta", "mu", "tau", "gamma1", "gamma2")))
```

Given this output, we can get the probability a player who is black
will win against and equally ranked player, when his current win ratio 
is exactly his average will be $1-\exp(-\alpha_j)$ while if they 
are playing as white their win probability will be
$1-\exp(-(\alpha_j + \gamma_1))$, so $\gamma_1$ being positive indicates the win
probability increases when playing white, which makes sense.
Playing white makes a player who has 50% chance of winning as black
have 54% chance of winning with white. Wikipedia says the difference is
4-6% so similar enough.

>>>>

knitr::opts_chunk$set(echo = FALSE, message = FALSE)

@@

Now we want to fit this multilevel logistic regression. Letting
$y_{ij}$ be the result of the $i$-th game by the $j$-th focal player,
then the model we will consider is

$$
P(y_{ij} = 1) = \frac{1}{1 + \exp(-(\alpha_j + \beta_j x_{ij} + \gamma z_{ij}))}
$$

where:

- $\alpha_j$ is a player level random effect
- $\beta_j$ is a player level random effect, accounting for the win ratio $x_{ij}$
- $\gamma$ is the coefficient of game level coefficients

We then partially pool the $\alpha$ and $\beta$ coefficients, so

$$
\alpha_j \sim \mathcal{N}(\mu_1, \tau_1)
$$
$$
\mu_1 \sim \mathcal{N}(0, 5)
$$
$$
\tau_1 \sim \mathcal{Cauchy}(0, 5)
$$

and similarly

$$
\beta_j \sim \mathcal{N}(\mu_2, \tau_2)
$$
$$
\mu_2 \sim \mathcal{N}(0, 5)
$$
$$
\tau_2 \sim \mathcal{Cauchy}(0, 5)
$$

We rescale the win ratio $x_{ij}$, by the average win ration of
player $j$. This means $\beta_j$ captures how a players win
probability changes as their historic performance deviates from 
the overall average win ratio.

For $z_{ij}$ we incorporate 2 covariates, namely the colour of 
the focal player and the ELO difference between the focal 
player and their opponent. So that means we have 
$\gamma_1$, which tells us how the win probability changes in going
from black to white,
and $\gamma_2$ gives how the win probability changes as the focal player
has a better ELO than their opponent.

```{r fit this model in stan}
#| eval: false

stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

stan_file <- "model3.stan"

fit3_ave$save_object("init_model_n10.RDS")
```

Some of the output of this model is shown below, namely the empirical posteriors
for each of the parameters.

```{r load in saved fit}
#| message: false

fit3_ave <- readRDS("init_model_n10.RDS")

mcmc_hist(fit3_ave$draws(c("alpha", "mu2", "tau2")))
mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))

rhat(fit3_ave)

#### Interpreting this output

Given this output, we can get the probability a player who is black
will win against an equally ranked player, when their current win ratio 
is exactly their average will be $\frac{1}{1+\exp(-\alpha_j)}$.
So a positive $\alpha_j$ indicates a player favoured to win 
even when playing black. The overall $\mu_1$ tells us about the 
global probability of winning as black against an equally ranked
opponent.
If a player
is playing as white their win probability will be
$\frac{1}{1+\exp(-(\alpha_j + \gamma_1))}$,
so $\gamma_1$ being positive indicates the win
probability increases when playing white, which makes sense.
Playing white makes a player who has 50% chance of winning as black
have approximately a 54% chance of winning with white. Wikipedia says the increase 
in win probability from playing white has been estimated to be 
4-6% so quite similar.

Similarly, $\gamma_2$ tells us about how the ELO of the focal player 
being larger than their opponent influences their win probability.
The estimate here indicates that a large ELO difference makes a player very
likely to win, as would be expected. For example, a player who
has a 50% chance of winning against an opponent with the same ELO
will have about a 56% of winning against an opponent with an ELO score
100 points lower.

### Varying History

We can refit this using varying numbers of games for the
current win ratio.

```{r 5 prev games}
#| eval: false

tidy_games <- map_dfr(top_players, get_hist, bullet_60, prev_n = 5) %>% 
  as_tibble()

stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

stan_file <- "model3.stan"

fit3_ave$save_object("init_model_n5.RDS")

```{r}
fit3_ave <- readRDS("init_model_n5.RDS")

mcmc_hist(fit3_ave$draws(c("beta", "gamma1", "gamma2")))
```

```{r 20 prev games}
#| eval: false

tidy_games <- map_dfr(top_players, get_hist, bullet_60, prev_n = 20) %>% 
  as_tibble()

stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

stan_file <- "model3.stan"

## this took 3 minutes

fit3_ave$save_object("init_model_n20.RDS")

```{r}

fit3_ave <- readRDS("init_model_n20.RDS")

mcmc_hist(fit3_ave$draws(c("beta", "mu1","gamma1", "gamma2")))
```

```{r prev 50 games}
#| eval: false

tidy_games <- map_dfr(top_players, get_hist, bullet_60, prev_n = 50) %>% 
  as_tibble()

stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

stan_file <- "model3.stan"

fit3_ave$save_object("init_model_n50.RDS")
```

```{r}
#| message: false
fit3_ave <- readRDS("init_model_n50.RDS")

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "gamma1", "gamma2")))

Using a longer history seems to give a model which fits quicker and
is more stable. But I'm not sure why that should be the case?

The size of the winner effects does appear to decrease as you look
at a longer amount of previous games. Unlikely to play more than 
50 games in a row maybe? Still clear positive effects even 
using 50 games.

## Comments

- There are some potential concerns with this Stan model and that it could
be made more efficient and better parameterized. These seem to be a
larger issue when the number of previous games used for the win ratio
is smaller, which would make sense as these would depart more from the mean.

- The size of these winner effects decreases as we look at their win ratio over
more historic games.

- Are there any other variables we should be controlling for here?

@@

@@

@@

@@

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))
mcmc_hist(fit3_ave$draws(c("alpha", "mu2", "tau2")))
```
